---
title: "JHU Covid 19"
author: "Riley Moore"
date: "2025-06-23"
output: html_document
---

## Included Libraries


```{r setup, message = FALSE}
library(tidyverse)
library(lubridate)

```

## Data input

The urls below are from John Hopkins Github, these 5 links provided all the data for this R markdown.

```{r get_jhu_data}

url_in <- "https://raw.githubusercontent.com/CSSEGISandData/COVID-19/master/csse_covid_19_data/csse_covid_19_time_series/"

file_names <- c("time_series_covid19_confirmed_global.csv",
    "time_series_covid19_deaths_global.csv",
    "time_series_covid19_confirmed_US.csv",
    "time_series_covid19_deaths_US.csv")
urls <- str_c(url_in,file_names)

uid_look_url <-"https://raw.githubusercontent.com/CSSEGISandData/COVID-19/master/csse_covid_19_data/UID_ISO_FIPS_LookUp_Table.csv"
```

## Variables for the Analysis 
The 4 variables we begin are read in from their respective file, global_cases, global_deaths, us_cases, and us_deaths

``` {r import_data, message = FALSE}

global_cases <- read_csv(urls[1])
global_deaths <- read_csv(urls[2])
us_cases <- read_csv(urls[3])
us_deaths <- read_csv(urls[4])

uid <- read_csv(uid_look_url) %>%
  select(-c(Lat, Long_, Combined_Key, code3, iso2, iso3, Admin2))

```
## Tidying up the Data

We tidy up the data by removing Latitude (Lat) and Longitude(Long_) from the data set as these are not important for our analysis.\
We also join together the data sets of global_case and global_deaths into one data set "global". We also read in the population for the global set from our 5th URL (uid). \
We do a similar joining together of the data sets us_cases and us_deaths into one data set US. We also filter both combined data set to only show data in which cases are over 0.\
```{r tidy_global_data, message = FALSE}

global_cases <- global_cases %>%
  pivot_longer(cols = 
                 -c('Province/State',
                    'Country/Region','Lat','Long',),
               names_to = "date",
               values_to = "cases") %>%
  select(-c(Lat, Long))

global_deaths <- global_deaths %>%
  pivot_longer(cols = 
                 -c('Province/State',
                    'Country/Region',
                    'Lat',
                    'Long',),
               names_to = "date",
               values_to = "deaths") %>%
  select(-c(Lat, Long))

global <- global_cases %>%
  full_join(global_deaths) %>%
  rename(Country_Region = `Country/Region`,
         Province_State = `Province/State`) %>%
  mutate(date = mdy(date))

global <- global %>% filter(cases > 0)

global <- global %>% 
  unite("Combined_Key",
        c(Province_State, Country_Region),
        sep = ", ",
        na.rm = TRUE,
        remove = FALSE)

global <- global %>% 
  left_join(uid, by = c("Province_State", "Country_Region")) %>% 
  select(-c(UID, FIPS)) %>%
  select(Province_State, Country_Region, date, cases, deaths, Population, Combined_Key)

```
```{r tidy_US_data, message = FALSE}
us_cases <- us_cases %>% 
  pivot_longer(cols = -(UID:Combined_Key),
               names_to = "date",
               values_to = "cases") %>%
  select(Admin2:cases) %>%
  mutate(date = mdy(date)) %>%
  select(-c(Lat, Long_))

us_deaths <- us_deaths %>% 
  pivot_longer(cols = -(UID:Population),
               names_to = "date",
               values_to = "deaths") %>%
  select(Admin2:deaths) %>%
  mutate(date = mdy(date)) %>%
  select(-c(Lat, Long_))

US <- us_cases %>% 
  full_join(us_deaths)

US <- US %>% filter(cases > 0)
```

# Visualizing the Data

We group our data by Country_Region to make it a little more meaniful. This way we can analyis covid by State for US cases and by country for global cases. \

A few new variables are made through simple computation, these are deaths_per_mill, cases_per_thou
```{r visualize_global_data, message = FALSE}
global_by_country <- global %>%
  group_by(Province_State, 
           Country_Region, 
           date) %>%
  summarize(cases = sum(cases), 
            deaths = sum(deaths), 
            Population = sum(Population)) %>%
  mutate(deaths_per_mill = deaths * 1000000 / Population) %>%
  select(Province_State, 
         Country_Region, 
         date, 
         cases, 
         deaths, 
         deaths_per_mill, 
         Population) %>% 
  ungroup()

global_country_totals <- global_by_country %>%
  group_by(Country_Region) %>%
  summarize(deaths = max(deaths), 
            cases = max(cases), 
            population = max(Population),
            cases_per_thou = 1000 * cases/population,
deaths_per_thou = 1000* deaths / population) %>%
  filter(cases > 0, population > 0)

global_country_totals %>% 
  slice_max(deaths_per_thou, n = 10) %>%
  select(deaths_per_thou, cases_per_thou, everything())

global_country_totals %>% 
  slice_min(deaths_per_thou, n = 10) %>%
  select(deaths_per_thou, cases_per_thou, everything())


```

```{r visulize_us_data, message = FALSE}

US_by_state <- US %>%
  group_by(Province_State, 
           Country_Region, 
           date) %>%
  summarize(cases = sum(cases), 
            deaths = sum(deaths), 
            Population = sum(Population)) %>%
  mutate(deaths_per_mill = deaths * 1000000 / Population) %>%
  select(Province_State, 
         Country_Region, 
         date, 
         cases, 
         deaths, 
         deaths_per_mill, 
         Population) %>% 
  ungroup()

US_totals <- US_by_state %>% 
  group_by(Country_Region, 
           date) %>%
  summarize(cases = sum(cases), 
            deaths = sum(deaths), 
            Population = sum(Population)) %>%
  mutate (deaths_per_mill = deaths * 1000000 / Population) %>%
  select(Country_Region, 
         date, 
         cases, 
         deaths, 
         deaths_per_mill, 
         Population) %>%
  ungroup()

US_by_state <- US_by_state %>%
  mutate(new_cases = cases - lag(cases),
         new_deaths = deaths - lag(deaths))
US_totals <- US_totals %>%
  mutate(new_cases = cases - lag(cases),
         new_deaths = deaths - lag(deaths))


```




## Plotting the Data
There are only plots for US data./
There are two US total plots one that show total cases against total deaths and one that shows total new_cases against total new_deaths//
There is one more data plot that is showing the new_cases against new_deaths in the State of California. 
```{r US_data_plots, warnings = FALSE, message = FALSE}

US_totals %>% 
  ggplot(aes(x = date, y = cases)) + 
  geom_line(aes(color = "cases")) + geom_point(aes(color = "cases")) +
  geom_line(aes(y = deaths, color = "deaths")) +
  geom_point(aes(y = deaths, color = "deaths")) + scale_y_log10() + 
  theme(legend.position = "bottom",
        axis.text.x = element_text(angle = 90)) +
  labs(title = "COVID 19 in US", y = NULL)

US_totals %>% 
  ggplot(aes(x = date, y = new_cases)) + 
  geom_line(aes(color = "new_cases")) + geom_point(aes(color = "new_cases")) +
  geom_line(aes(y = new_deaths, color = "new_deaths")) +
  geom_point(aes(y = new_deaths, color = "new_deaths")) + scale_y_log10() + 
  theme(legend.position = "bottom",
        axis.text.x = element_text(angle = 90)) +
  labs(title = "COVID 19 in US", y = NULL)

state <- "California"
US_by_state %>%
  filter(Province_State == state) %>%
  ggplot(aes(x = date, y = new_cases)) + 
  geom_line(aes(color = "new_cases")) + geom_point(aes(color = "new_cases")) +
  geom_line(aes(y = new_deaths, color = "new_deaths")) +
  geom_point(aes(y = new_deaths, color = "new_deaths")) + scale_y_log10() + 
  theme(legend.position = "bottom",
        axis.text.x = element_text(angle = 90)) +
  labs(title = str_c("COVID 19 in ", state), y = NULL)

```

## Model and Predictions
Here is a linear model for predicting cases on a Global scale grouped by Countries/
```{r global_model_pred}
mod <- lm(deaths_per_thou ~ cases_per_thou, data = global_country_totals)

global_country_totals_pred <-global_country_totals %>% 
  mutate(pred = predict(mod))



```

## Model Plots

Here is a plot that is showing the Predicted deaths per thousand vs the actual death per thousand.

```{r global_pred_plot}
global_country_totals_pred %>%
  ggplot() +
  geom_point(aes(x = cases_per_thou, y = deaths_per_thou), color = "blue") +
  geom_point(aes(x = cases_per_thou, y = pred), color = "red")
```

## Bias

In this data there are a lot of different bias./ The biggest bias is the varying level of healthcare available in different countries and different states. Futhermore each country had different systems in place to mitagate the exposure of covid.